---
title: "LOOE"
author: "Dolors Pelegrí"
date: "`r Sys.Date()`"
#output: 
#  rmarkdown::html_vignette:
#    theme: Architect:
#    toc_float: true
output:
  BiocStyle::html_document:
    toc_float: true
    
pages:
  markdown:
    extra: true
    
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


# Overview



# Matrix Multiplication

## Simple matrix multiplication (withou blocks) :

If you want to perform a simple sequential matrix multiplication you can call the function `blockmult()` with DelayedArray data type or any other R data type. In the example bellow we perform a product with A(500x500) and B(500x750).

```{r warning=FALSE, message=FALSE  }
library(BigDataStatMeth)
library(DelayedArray)
n <- 500
p <- 750

A <- matrix(rnorm(n*n), nrow=n, ncol=n)
B <- matrix(rnorm(n*p), nrow=n, ncol=p)
DA <- DelayedArray(A) # Get A as Delayed Array
DB <- DelayedArray(B) # Get B as Delayed Array

AxB <- blockmult(A, B)
AxBDelay <- blockmult(DA, DB)

AxBDelay [1:5,1:5]

all.equal(AxB, AxBDelay)
all.equal(A%*%B, AxBDelay)

```

you can optimize the performance making this calculus in parallel with `paral=true` .  

```{r warning=FALSE, message=FALSE  }

AxB <- blockmult(A, B, paral = TRUE)
AxBDelay <- blockmult(DA, DB, paral = TRUE) # using Delayed Array data type

all.equal(AxB, AxBDelay)

```


## Block matrix multiplication

A block matrix or a partitioned matrix is a matrix that is interpreted as having been broken into sections called blocks or submatrices. Intuitively, a matrix interpreted as a block matrix can be visualized as the original matrix with a collection of horizontal and vertical lines,
which break it up, or partition it, into a collection of smaller matrices [(more information on block matrix multiplication)](https://en.wikipedia.org/wiki/Block_matrix).

The code bellow performs a matrix multiplication with A(500x500) and B(500x750) with (10x10) blocks. 

```{r warning=FALSE, message=FALSE  }

AxB <- blockmult(A, B, block_size = 10)
AxBDelay <- blockmult(DA, DB, block_size = 10 )

all.equal(AxBDelay,A%*%B)
all.equal(AxB, AxBDelay)


```

if block_size > cols(A) and rows(A) and cols(B) and rows(B) then blocks_size is set to min( ols(A), rows(A), cols(B), rows(B)) 

Like in simple matrix multiplication you can make operations in parallel with `paral = TRUE` .

```{r warning=FALSE, message=FALSE  }

AxB <- blockmult(A, B, block_size = 10, paral = TRUE)
AxBDelay <- blockmult(DA, DB, block_size = 10, paral = TRUE )

all.equal(AxBDelay,A%*%B)
all.equal(AxB, AxBDelay)

```


# Cross-product and Transposed Cross-product

To perform a cross-product $C = A^t A$ you can call `bdcrossprod()`.

```{r warning=FALSE, message=FALSE  }
n <- 500
m <- 250
A <- matrix(rnorm(n*m), nrow=n, ncol=m)
DA <- DelayedArray(A)

# Cross Product
cpA <- bdcrossprod(A, transposed = FALSE)  
cpDA <- bdcrossprod(DA)  # With DelayedArray data type and no transposed parameter

all.equal(cpDA, crossprod(A))
```

you may also set `transposed=TRUE` (default value transposed=false) to perform a Transposed Cross-product $C = A A^t$

```{r warning=FALSE, message=FALSE  }
# Transposed Cross Product
tcpA <- bdcrossprod(A, transposed = TRUE)
tcpDA <- bdcrossprod(DA, transposed = TRUE) # With DelayeArray data type

all.equal(tcpDA, tcrossprod(A))

```

# Matrix Vector Multiplication 

You can perform a weighted cross-product $C = X^ t w X$ with `bdwcrossprod()` given a matrix X as argument and a vector or matrix of weights, w . 

## Weighted Cross-product and Weighted Transposed Cross-product

```{r warning=FALSE, message=FALSE  }
n <- 250
X <- matrix(rnorm(n*n), nrow=n, ncol=n)
u <- runif(n)
w <- u * (1 - u)
DX <- DelayedArray(X)
Dw <- DelayedArray(as.matrix(w))
  
wcpX <- bdwproduct(X, w,"xwxt")
wcpDX <- bdwproduct(DX, Dw,"xwxt") # with DelayedArray

wcpDX[1:5,1:5]

all.equal(wcpDX, X%*%diag(w)%*%t(X))

```

## Weighted Transposed Cross Product 

With argument `transposed=TRUE`, we can perform a transposed weighted cross-product $C = A w A^t$  

```{r warning=FALSE, message=FALSE  }

wtcpX <- bdwproduct(X, w,"xtwx")
wtcpDX <- bdwproduct(DX, Dw,"xtwx") # with DelayedArray

wtcpDX[1:5,1:5]

all.equal(wtcpDX, t(X)%*%diag(w)%*%X)

```

# Inverse Cholesky
<p> The Cholesky factorization is widely used for solving a system of linear equations whose coefficient matrix is symmetric and positive definite.</p>
<p style="text-align: center;">$A = LL^t = U^tU$ </p>
where $L$ is a lower triangular matrix and U is an upper triangular matrix.
To get the Inverse Cholesky we can use the function `bdInvCholesky()`


```{r warning=FALSE, message=FALSE  }

# Generate a positive definite matrix
Posdef <- function (n, ev = runif(n, 0, 10)) 
{
  Z <- matrix(ncol=n, rnorm(n^2))
  decomp <- qr(Z)
  Q <- qr.Q(decomp) 
  R <- qr.R(decomp)
  d <- diag(R)
  ph <- d / abs(d)
  O <- Q %*% diag(ph)
  Z <- t(O) %*% diag(ev) %*% O
  return(Z)
}

A <- Posdef(n=500, ev=1:500)
DA <- DelayedArray(A)

invchol <- bdInvCholesky(A)
Dinvchol <- bdInvCholesky(DA)

round(invchol[1:5,1:5],8)

all.equal(Dinvchol, solve(A))

```

# SVD Decomposition

## svd
## Jacobi decomposition




# LOOE

